Do convolutional neural networks exhibit human-like attentional effects? Conflicting results from a visual search task

### 2019-05-27 docs/ccneuro2019/paper.tex
rewrite ccneuro paper abstract

\section{Abstract}
{
\bf
In machine learning and artificial intelligence, attention mechanisms are proposed to 
improve classification accuracy. In cognitive science, however, attention mechanisms are 
often defined in terms of experimental manipulations that impair performance. For example,
psychologists have long used a now classic form of visual search task to determine what 
factors increase reaction times or decrease accuracy. Here we investigate the ability of 
neural networks to perform the classic visual search task, with the goal of reconciling 
what appear to be different definitions of attention in cognitive science and in 
artificial intelligence. We first show that convolutional neural networks (CNNs) trained 
for image classification can show human-like attentional effects, replicating and 
extending a previous study. Then we demonstrate that, by changing how we train the 
networks and the statistics of the training data, these effects can largely be 
eliminated, i.e., that CNNs are capable of performing the task without explicit attention 
mechanisms. We go on to show that manipulating the discriminability of the target and 
distractors in the task can produce attentional effects in spite of the changes we made to
training. We discuss the implications of these findings for the study of attention.
}
\begin{quote}
\small
\textbf{Keywords:} 
attention;visual search
\end{quote}

>>>>

\section{Abstract}
{
\bf
Selective attention mechanisms preferentially route some sensory inputs for further 
processing, while necessarily discarding others. Researchers studying visual
attention have often divided selective attention into overt and covert mechanisms. 
Overt mechanisms involve orienting the eyes and/or head, while covert mechanisms 
occur "centrally", in the brain, without overt orientation. Cognitive scientists have long
used a new classic form of visual search task to investigate theoretical mechanisms of 
selective covert visual attention. Here we build on previous studies that examine the 
parallels between the visual system and convolutional neural networks (CNNs), by testing 
whether CNNs exhibit human-like attentional effects while performing this visual search 
task. Following a typical approach, we first show that CNNs trained 
for image classification can show human-like attentional effects. Then we demonstrate 
that, by changing how we train the networks and the statistics of the training data, 
these effects can largely be eliminated, i.e., that CNNs are capable of performing the 
task without explicit attention mechanisms. We go on to show that manipulating the 
discriminability of the target and distractors in the task can produce attentional 
effects in spite of the changes we made to training. Based on these results, we conclude 
that CNNs do not show attentional effects.
}
\begin{quote}
\small
\textbf{Keywords:} 
attention;visual search
\end{quote}
